A high-level programming framework to support parallel processing of communication intensive \\ stream applications

### 2013-03-18 paper.tex
Re commit

% *** SPECIALIZED LIST PACKAGES ***
%
%\usepackage{algorithmic}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

@@

%\usepackage{mdwmath}
%\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/

@@

% *** SUBFIGURE PACKAGES ***
%\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}
%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

@@

@@

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Title goes here}

@@

\begin{abstract}
%\boldmath
The abstract goes here.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

The constant advancement in server hardware has made systems with large number of hardware threads[] readily available. In addition to this the presence of co-processors[], network processors[], Graphics Processing Units ( GPUs ) \& General Purpose GPUs ( GPGPUs ) has dramatically improved the capability of commodity hardware. These developments has paved way to a platform with immense computing potential which existing applications already take advantage of. Along with the improvements in the Multi-Core \& Many-Core processor architecture the memory and I/O speeds have also seen significant improvement. In particular the network subsystem has seen significant growth, in combination with the improvements to optical communication it has lead to 10GbE interfaces being quite common in enterprise servers. While 40GbE and 100GbE are available for backbone infrastructure but would eventually follow into commodity hardware.

Network applications represent those applications that operate on any of the seven network layers described in the OSI Model. \{describe more here\}

Several previous work have discussed the process of parallelizing network applications[][][] to enable it to run faster in such multi threaded systems. The development of Software routers[][][] is encouraged further because it costs only a fraction of the commodity hardware routers. Moreover the advancements in the network interfaces has made faster interfaces readily in combination with the multi core servers the performance  It has been shown that such routers can be used to process packets at line rate [][][] which has been a challenge on its own. Network applications including those that operate on few or all of the network layers have been designed with sequential processing as the target environment. The shift to multi core architectures in recent times has induced such applications to utilize parallel processing as a alternate means of computing. While this could lead to significant improvement in performance it also suffers from the same problems as any parallel application. These include scalability, load-balancing, data dependency and the presence of non parallel regions in the program apart from those that might arise due to that specific application. The scaling of software routers has been discussed extensively in RouteBricks[] which shows the problems suffered by software routers when they are scaled beyond a single system. One could argue that complexity of network applications such as Deep Packet Inspection ( DPI ), Packet Forwarding, etc., could be handled by a single many socket Multi-Core SMP system. However, as the complexity of network applications develops it would be inevitable but to move to a multi node environment in order to perform the computations and to handle more network bandwidth.

Large scale computing clusters has been used in high-performance scientific application to solve problems and applications of very large size. It uses a cluster of compute nodes and distributes the workload across them in order to solve it faster. These systems utilize parallel programs consisting of both threads and MPI to break massive problems into smaller workload that can be executed in parallel. In addition, the presence of GPUs and GPGPUs requires a mixture of parallelization constructs such as OpenCL, CUDA, etc., in order to achieve the best performance in the system. In this paper we discuss how some of the techniques used by these applications could be used to parallelize network applications.

Click Modular Software Router[] consists of extensive elements which are used to construct software routers. Click primarily targets packet processing applications especially those that operate on L4 - L7. \{elaborate more here\}

Discuss about what each sections contains here.....

@@

Talk about click...\\

Describe about what would be the best way to parallelize using threads and MPI.. provide reference.. \\

\section{Design}

This section describes the design principles related to our methodology. The objective is to determine the requirements necessary to parallelize network applications. Our methods focus on ways to improve the click system in some cases even extend its functionality.

\subsection{Emphasize on user level execution environment}

The existing software routers are predominantly implemented in the kernel space for the ease of access to resources and can bypass the OS overhead incurred when data is accessed through the network subsystem. While this alternative provides better performance it also leads to concerns regarding stability and security of the system. Newer techniques such as Netmap[] allows direct access to network buffers which makes it possible to write high performance packet processing applications in userspace itself. Typically network applications which operate on L4 - L7 are developed in the kernel space in order to intercept the network packets before it flows through the OS. Whereas those applications that utilize L1 - L3 are present in the user space. The objective here is to completely switch to user space to solve any of the stability or security concerns. In addition, scenarios where the applications that operate across most or all the layers could be improved by combining the operations across the different layers in user space itself.

\subsection{Eliminate the overhead associated with creation of dynamic objects}

Click constructs application graphs based on dynamic creation of objects and passing the data packets across the virtual functions. Though this is a very elegant approach in creating the software router it does incur the penalty of the virtual function tables. The click-tool[] addresses this issue but it does not solve the problem entirely. The construction of the software router at run time is the primary reason behind it. This mainly arises due to the presence of kernel module which installs the router as a kernel module based on the input configuration. However in the user level it would not be necessary for a separate process to initialize the router instead we could generate the entire program statically including the necessary flow associated with the specific application. This not only eliminates the overhead associated but also opens up means to optimize the graph further.

\subsection{Optimize based on application graph}

Network applications especially that operate on the lower layers are very sensitive to the overhead associated with operating on the data separately on each layer. Instead given an application graph if we were to provide a sequential flow of the different functions at the various layers then it would be possible for the compiler to optimize the regions of program into a single and optimized block of code. This would also improve the cache locality of the data accessed by consecutive functions. This also presents us with the ability to perform other types of stream graph optimizations[] which would not be possible if the specific graph was constructed dynamically at run time.

\subsection{Automated data parallelism}

Network traffic is a classic example by which data parallelism can be utilized to achieve scalable parallelism. Click's support for parallelism in user level is described as experimental, it is implemented by using a task graph consisting of each work associated with each element which is executed in the threads. The support for parallelism in the kernel mode is better in comparison to that of user mode. Since our target environment is in the user space the proposed design involves a different approach to which we can extract data parallelism. We first split the data across multiple queues and then replicate the application graph across to process the data the queues in parallel. An important condition here is to make sure packets of the same flow should pass through the same queue. Since the application graph can be replicated it also makes it possible to dynamically load balance several applications based on the workload.

\subsection{Efficient implementation of task parallelism}

In generating the application graph as a sequence of operations it also makes it possible to process different stages of the operation in a pipelined manner. This when combined with stream optimizations and could be used to determine which steps of the graph is best suited to form the stages of a pipeline. The condition to balance here is the weight associated with computation in that stage with that of the overhead incurred in computing the data in two separate stages. It would be necessary to determine whether the additional over head involved in executing the task in parallel provides any significant improvement in performance.

\section{Implementation}

Click was designed as a means to easily implement a software router on top of a Linux OS. This is implemented in two different ways one is by installing a kernel module and the other is a program that executes in the user level. Both of them accept a configuration file described in click's language specification. This is then used to construct the router by linking the virtual functions for the push and pull operations of the different elements in accordance with the specified graph.

\subsection{Modifications to click system}

\subsubsection{Static Generation Of ???}
The first step in the design was to overcome the limitation of using click's infrastructure to construct the software router which consists of dynamically initialized objects and virtual functions. This was necessary due to two important conditions, one was to eliminate the overhead associated with the dynamic construction of the objects and virtual functions and the other was to provide the compiler a chance to optimize the program during the compilation phase. This would however result in not being able to construct the software router using click's infrastructure but would require the application specific code to be recompiled. This could be a significant hurdle in the case of running the software router in kernel mode where the router configuration is installed by the click kernel module in which case the entire kernel module should be recompiled and reinitialized. Whereas in user space this would not be an issue since routers of different configurations are basically treated as separate applications and would just have to go through an additional compilation phase before being used. The added advantage of this would be to enable network applications that operate on L1 - L3 layers to use the packet processing ( L4 - L7 ) infrastructure provided by click elements.

\subsubsection{Implementation of data parallelism}

--Talk about MPI \& threads\\

--Our goal here is to focus on the network application as a workload and characterize it as such without involving any kind of overhead that might occur due to the network.\\

>>>>

\usepackage{listings}
\usepackage{color}

\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}

\lstset{ %
  language=C,                % the language of the code
  basicstyle=\footnotesize\ttfamily,           % the size of the fonts that are used for the code
  numbers=right,                   % where to put the line-numbers
  numberstyle=\tiny\color{gray},  % the style that is used for the line-numbers
  stepnumber=0,                   % the step between two line-numbers. If it's 1, each line 
                                  % will be numbered
  numbersep=0pt,                  % how far the line-numbers are from the code
  %backgroundcolor=\color{white},      % choose the background color. You must add \usepackage{color}
  showspaces=false,               % show spaces adding particular underscores
  showstringspaces=false,         % underline spaces within strings
  showtabs=false,                 % show tabs within strings adding particular underscores
  %frame=single,                   % adds a frame around the code
  rulecolor=\color{black},        % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. commens (green here))
  tabsize=2,                      % sets default tabsize to 2 spaces
  captionpos=b,                   % sets the caption-position to bottom
  breaklines=true,                % sets automatic line breaking
  breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
  title=\lstname,                   % show the filename of files included with \lstinputlisting;
                                  % also try caption instead of title
  keywordstyle=\bfseries,          % keyword style
  commentstyle=\color{dkgreen},       % comment style
  stringstyle=\color{mauve},         % string literal style
  escapeinside={\%*}{*)},            % if you want to add LaTeX within your code
  morekeywords={*,...}               % if you want to add more keywords to the set
}

% *** SPECIALIZED LIST PACKAGES ***
%
\usepackage{algpseudocode}
\usepackage{algorithm}
%\usepackage{algorithmic}
%\usepackage{algorithmicx}
% algorithmic.sty was written by Peter Williams and Rogerio Brito.
% This package provides an algorithmic environment fo describing algorithms.
% You can use the algorithmic environment in-text or within a figure
% environment to provide for a floating algorithm. Do NOT use the algorithm
% floating environment provided by algorithm.sty (by the same authors) or
% algorithm2e.sty (by Christophe Fiorio) as IEEE does not use dedicated
% algorithm float types and packages that provide these will not provide
% correct IEEE style captions. The latest version and documentation of
% algorithmic.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithms/
% There is also a support site at:
% http://algorithms.berlios.de/index.html
% Also of interest may be the (relatively newer and more customizable)
% algorithmicx.sty package by Szasz Janos:
% http://www.ctan.org/tex-archive/macros/latex/contrib/algorithmicx/

@@

\usepackage{mdwmath}
\usepackage{mdwtab}
% Also highly recommended is Mark Wooding's extremely powerful MDW tools,
% especially mdwmath.sty and mdwtab.sty which are used to format equations
% and tables, respectively. The MDWtools set is already installed on most
% LaTeX systems. The lastest version and documentation is available at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/mdwtools/

@@

% *** SUBFIGURE PACKAGES ***
\usepackage[tight,footnotesize]{subfigure}
% subfigure.sty was written by Steven Douglas Cochran. This package makes it
% easy to put subfigures in your figures. e.g., "Figure 1a and 1b". For IEEE
% work, it is a good idea to load it with the tight package option to reduce
% the amount of white space around the subfigures. subfigure.sty is already
% installed on most LaTeX systems. The latest version and documentation can
% be obtained at:
% http://www.ctan.org/tex-archive/obsolete/macros/latex/contrib/subfigure/
% subfigure.sty has been superceeded by subfig.sty.

\usepackage[font=small,center]{caption}
%\usepackage[font=footnotesize]{subfig}
% subfig.sty, also written by Steven Douglas Cochran, is the modern
% replacement for subfigure.sty. However, subfig.sty requires and
% automatically loads Axel Sommerfeldt's caption.sty which will override
% IEEEtran.cls handling of captions and this will result in nonIEEE style
% figure/table captions. To prevent this problem, be sure and preload
% caption.sty with its "caption=false" package option. This is will preserve
% IEEEtran.cls handing of captions. Version 1.3 (2005/06/28) and later 
% (recommended due to many improvements over 1.2) of subfig.sty supports
% the caption=false option directly:
%\usepackage[caption=false,font=footnotesize]{subfig}

%
% The latest version and documentation can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/subfig/
% The latest version and documentation of caption.sty can be obtained at:
% http://www.ctan.org/tex-archive/macros/latex/contrib/caption/

@@

\newcommand{\INDSTATE}[1][1]{\STATE\hspace{#1\algorithmicindent}}
\newcommand{\comment}[1]{}

@@

\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{A high-level programming framework to support parallel processing of communication intensive \\ stream applications}

@@

\begin{abstract}
%\boldmath
Stream applications are often limited in their performance by the communication system employed. In systems that use conventional 1GbE or 10GbE standard, the Operating System (OS) handles most if not all of the network operations. In such cases, the communication stack, which was not designed to handle such tremendous amounts of data, acts as a bottleneck and restricts the performance of the application. An optimum framework for computing and communicating in parallel is essential for the continued improvement in performance.

We propose a high-level programming framework that integrates the packet processing operations with the stream applications and provides a common interface to execute both the application and packet processing operations in parallel on the multicore processors.

We utilize the ability of network interface controller (NIC) to classify packets on to multiple hardware queues. We combine this with system software that allows direct access to the hardware queues. We develop an interface in our framework for application processes to read or write to in parallel. Our system integrates this with the application by a series of packet processing operations that transform the data from the application to packets suitable for transmission. Representing these tasks as stages in the application itself, we are able to parallelize them using techniques similar to those used for parallelizing regular computations.

All the operations required for the communication are performed in the application, bypassing the OS to send and receive data directly over the network interface. By overcoming the bottleneck induced by performing network operations in the OS and by parallelizing both the computations and packet operations together, we are able to boost the performance of the application.

\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the conference you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals/conferences frown on
% math in the abstract anyway.

@@

Stream applications are designed in a manner in which they operate on data that flows as a \textit{stream}. This data stream passes through different stages of computations, to be transformed from query to result \cite{Thies:2002:SLS:647478.727935} \cite{Halbwachs91thesynchronous} \cite{Stephens95asurvey}. This is different from conventional computation model where data is stored and then processed. By eliminating the need to store data and processing it on the go, it makes it possible for stream applications to compute on large amounts of data.

Stream applications such as file storage systems, database query management, data mining, etc., that operate on vast amount of data are often distributed over multiple systems. Traditional stream applications do target multicore processors for performing their computations in parallel. Due to this and the exponential rate at which multicore processors are growing the amount of computations performed by these stream applications would further increase and eventually lead to processing even larger amounts of data. With the growth in the complexity of these applications it is necessary to leverage on a multicore environment to improve their communication performance as well. Such stream applications that operate on large amounts of data at very high bandwidth are often limited in their performance by the communication stack of the OS kernel\cite{Wu:2007:PPB:1323954.1323957}\cite{Wu:2007:PAL:1227865.1228021}\cite{Dobrescu09routebricks:exploiting}\cite{Han:2010:PGS:1851275.1851207}\cite{Kohler2000}. In the continued race for improvement of multicore and many core architectures, it is becoming essential to improve the processing speed of packet operations along with those of the application in order to scale the performance of the application.

When performing packet processing operations for communication intensive stream applications, a major challenge is the very high levels of performance required. These applications process large numbers of data streams in real time.  If the communication system cannot keep up with the rate at which packets arrive due to these data streams, then they must be dropped or remain unprocessed.

In building a framework for such applications on general purpose computers, it may be possible to exploit two hardware features which are multicore processors and multi queue network controller that could allow dramatically improved performance. Using this it makes it possible to do much of the packet processing in parallel using the multiple cores that are available on modern processors.

A major bottleneck for packet processing on general purpose computers is the overhead of the operating system's network stack. In particular, for applications that run as user processes, data must be copied between OS kernel space and user space. Previous research use alternate mechanism to access packets from the network hardware directly due to this\cite{Dobrescu09routebricks:exploiting}\cite{Han:2010:PGS:1851275.1851207}\cite{Kohler2000}. Most of these use modern network hardware in combination with a particular systems software that allows direct access to hardware-controlled packet queues to dramatically reduce overheads and improve performance. However, constructing stream applications that use these features is difficult. Furthermore, many stream applications have a very specific communication pattern, typically operating on data between a source and sink in the form a \textit{data stream}. Data in a stream must be processed sequentially, whereas those from separate streams can often be processed in parallel.

Modern network hardware typically has a mechanism for separating arriving packets into multiple queues, where packets that belong to the same flow are guaranteed to be placed in the same queue\cite{micro2008} \cite{intel2010}. In many stream applications we can utilize this to operate on multiple streams of data by processing each of the queues in parallel. However, building such applications is complex because it requires explicit parallel programming, and it requires explicit interaction with the low-level network interfaces. Research focuses on improving the performance of network applications by parallelizing the packet processing tasks on multicore architectures \cite{Dobrescu09routebricks:exploiting} \cite{Han:2010:PGS:1851275.1851207}. We parallelize both the application and packet operations to gain significant improvement in performance.

\section{Contribution}

In this paper we propose a high-level programming framework for building communication intensive stream applications. We focus on integrating the packet processing and application tasks into a single graph and how we use this to extract data parallelism. The developer describes their program as a graph of \textit{actors}. Each actor consumes one or more \textit{tokens} of data, which may be simple data such as integers, or more complex data such as packets, arrays or objects and produces one or more tokens. The framework can be used to automatically map the application over multiple parallel cores, and interface with the \texttt{netmap} APIs \cite{Rizzo:2012:RNI:2090147.2103536} for accessing data through the network buffers.

The system can be divided into three main parts:

\begin{itemize} 
\item A subsystem for interfacing with the network hardware through the \texttt{netmap} APIs to send and receive data from the application in user-space without going through the operating system. (Section \ref{ppio}) 
\item Given a graph that represents the workload to be performed by the application, we construct and integrate the packet processing operations represented as a directed acyclic graph (DAG) to it. (Section \ref{ppe})
\item A run-time system that executes these tasks in parallel and explicitly schedules them on the multicore processor. (Section \ref{ppt})
\end{itemize}

Even though we integrate the system with the application the techniques for extracting parallelism from the application is not in scope of this paper. The application is considered to be in a streaming model and the tasks corresponding to the various functions are considered to be parallel in nature.

@@

Let us consider a sample application that operates on data streams. The important operations of the application are represented in the stream graph shown in figure \ref{fig1}. The application operates by performing encryption on the data that matches predefined rules like file names, access level of user, etc., otherwise it performs compression and store the information in the location specified. Considering a scenario in which, a group of servers are used to perform these operations by evenly distributing them over the different nodes.

\begin{figure}[ht]
\centering
\subfigure
{
	\includegraphics[width=3.5in]{ip-data-proc}
}

\subfigure
{
	\includegraphics[width=3in]{op-data-proc}
}
\caption{Sample stream application}
\label{fig1}
\end{figure}

The main challenge that arises in such a situation is how to tackle such large number of computations from the application and the network operations due to the intensive communication. Two important features enable us to tackle this challenge posed by such a system. Firstly, the operations of both the application and the network is parallel in nature and secondly, the presence of multicore processors to execute the application.

In our system we use an efficient parallel processing framework that could leverage on multicore processors and multi queue network controllers to achieve this. The performance of multicore processors has dramatically improved the capability of commodity hardware. With this we would be able to meet the computational demand of the application. In addition, the presence of multi queue network interface controller which was designed to handle multiple streams of data could help us achieve the necessary communication performance.

An important part of our system is the tasks related to packets such as Userdatagram Protocol (UDP), Internet Protocol (IP), etc., We utilize the click modular router for these packet processing tasks. Click modular router is a tool that includes a library of elements which is used in building flexible, re-configurable and modular software routers \cite{Kohler2000}. It offers a platform that is flexible for rapid development of newer network protocols and achieves good performance. An improved version known as SMP Click supports thread based parallel processing of network related tasks\cite{Chen:2001:FCP:647055.759948}. Until recently click's performance in the user space was not as good as its kernel space equivalent. This is mainly due to its extensive use of polling driver in the kernel space that was not possible in user space. Netmap I/O\cite{Rizzo:2012:RNI:2090147.2103536} provides a set of APIs that could directly access the packets from the NIC and it helped in the recent improvement of click's user space performance. We found the simpler construction of \texttt{netmap} easier to adapt to our framework compared to other user space packet access APIs \cite{Rizzo:2012:RNI:2090147.2103536} \cite{1564468} \cite{Krasnyansky}.

Section \ref{dppl} represents the design principles that we follow to conceptualize our framework. Section \ref{imple} describes the implementation details used in the construction of the system. In section \ref{ppe}, we show the challenges involved in using a system consisting of processes that are used to actually perform the work. The extent to which we extract parallelism ranges not only to the packet tasks but to the low level access of data I/O to the network interface hardware through certain additions to the \texttt{netmap} APIs. Finally in Section \ref{app-const} we present a simple application constructed from operations such as compression and encryption in order to evaluate our high-level framework.

\section{Design Principles}
\label{dppl}

The design of a high-level programming framework that integrates packet processing tasks into the application requires overcoming several challenges. \comment{The entire system is driven by the extent to which parallelism can be extracted. Some of the techniques that can be applied to the packet processing tasks can be used for the tasks related to the application.} In this section, we describe these challenges and the principles that are used to tackle them. \comment{We focus on the methodologies related to the packet processing tasks and the parallel processing engine itself.}

\subsection{Userspace Packet I/O}

For a packet processing system to be integrated with an application it is necessary to have an efficient mechanism for accessing the data from the NIC without interacting with the operating system. Rapid data access here is crucial for the extraction of data parallelism at later stages. \comment{In using the techniques described below we exploit the modern NIC architecture to its full potential.}

\subsubsection{Handling Multiple Hardware Queues}

The development of virtualization and improved network flow handling has lead to hardware level multi queue support in the NICs. During transmission concurrent writes to different queues are possible, enabling multiple processes to send data simultaneously. During reception the NIC classifies the packet on to one of the receive queues using a technique known as Receive Side Scaling (RSS)\cite{micro2008} \cite{intel2010}. It uses the header information or tuples to classify packets on to different hardware queues. Multi queue support can handle several streams of data in parallel and it is exploited extensively in our framework to process data concurrently. By offloading the classification of packets to the NIC, it removes any overhead associated with packet classification in the application. We use these hardware queues as a means of not only accessing the data in parallel but balancing the workload across different threads. This is done by controlling the number of queues that each process handles based on the rate at which packets are received in each of the queues. It is necessary to have different processes accessing the transmit and receive queues to maximize concurrent access to data.

\subsubsection{Flow based Packet Batching}

Packet batching is a common method used to boost the performance of network applications. Packet batching combines several packets and performing operations on them together, thus reducing the overhead involved in the I/O operations.

In our framework we utilize batching but the number of packets that forms a batch is not fixed but varies based on the available space in the hardware queue. This number reflects the amount of data we are able to process before issuing hardware synchronization calls and forms a important means by which we reduce the usage of system calls.

\subsubsection{Reduction of Per Packet System Calls}

System calls are required in order to synchronize data between the NIC hardware and the buffer memory used for intermediate storage of packets. Previous work describe the overhead associated with per packet system calls and provide means of optimizing this\cite{1564468}\cite{Han:2010:PGS:1851275.1851207}\cite{Rizzo:2012:RNI:2090147.2103536}.

We employ an \textit{adaptive scheme} described in section \ref{ppio} that uses the above principle. Our framework balances the system calls based on the communication rate and application processing speed. The transmission and reception are handled differently to optimize the usage of system calls in each of the case independently. During transmission we issue system calls to synchronize the available buffer space based on the rate at which we transmit data. This prevents degradation in latency for larger packets and improves performance for smaller packets given the same bandwidth at which we transmit in both the cases. During reception of packets the system reads packets at the rate at which it is able to process them.

\subsection{Packet Processing Graph}

\comment{Understanding the set of packet processing tasks associated with the application is essential to performing them in parallel.} We use a graph based representation of packet processing tasks to determine the dependency involved for the different stages and determine the relation with that of the different tasks of the application. The graph contains the stages through which the data, associated with the application, is transformed into network packets and vice versa. It shows which of the tasks can be performed in parallel with those of the application and highlights those that have data dependency on a specific task in the application. This graph also enables optimizations based on the characteristics of the application or the architecture used for execution. Figure \ref{fig2} shows a User Datagram protocol (UDP) packet generation represented in this graph.\\

\begin{figure}[ht]
\centering
\includegraphics[width=3.5in]{pack-op}
\caption{Simplified udp packet construction}
\label{fig2}
\end{figure}

\subsection{Application Task Graph}

\comment{Application task graph represents the operations associated in performing the computations.} The subset of the application that we are interested in parallelizing is the stream processing component. In such components the communication between tasks is predictable and there is correlation between the stages of the application. Both the packet processing and application tasks can be represent by a directed acyclic graph (DAG) similar to the packet processing graph. The packet processing tasks operate on packets and the application tasks operate on the data. Since we focus more on parallel packet processing, the optimization in relation to the application graph are not discussed. It is assumed that a static version of the application graph is available and we focus more on executing these tasks in parallel with that of the packet processing tasks.

\subsection{Integration}

The sequence of tasks that can be executed in parallel is determined at this stage. The integration stage is used to combine the application graph and the packet processing graph into a larger graph that unifies and represents the overall operations. The dependency relations determined on the previous cases are used in establishing links between the graphs. We determine the sequence of tasks that can be executed in parallel from both the application task graph and packet processing graph. The combined set of tasks is then used by the parallel processing engine.

\subsection{Parallel Processing Engine}

Two important aspects that the parallel processing engine must tackle is the scheduling of tasks and balancing the load across the available resources. The parallel processing engine consists of processes which are bound to specific core and act as engines to execute the workload. It is constructed in a manner in which its flexible and can adapt to utilize the resources efficiently. The tasks are scheduled across these execution engines based on the available resources and its requirements. In this manner the workload is distributed uniformly while retaining control over the different execution elements.

\section{Implementation}
\label{imple}

Most of the existing stream applications depend on the operating system for most of their communication tasks. In commodity systems, the interaction with the operating system for network operations is a necessary abstraction to handle multiple applications. However, in the case of specialized servers which cater to specific applications it would be beneficial to interact directly with the network device. Moreover, in the case of intensive network traffic the operating system could act as a bottleneck due to the overhead associated with data flowing through the kernel before being transmitted through the network device\cite{Wu:2007:PPB:1323954.1323957}\cite{Wu:2007:PAL:1227865.1228021}. Eliminating this would be possible by either executing the application in the kernel space which could compromise stability of the system or by using an interface for accessing the NIC buffers from the user space itself. The later case of accessing the buffers of network hardware in user space is supported by several interfaces such as \texttt{netmap} \cite{Rizzo:2012:RNI:2090147.2103536}, PF\_RING DNA \cite{1564468}, UIO-IXGBE \cite{Krasnyansky}, etc., Even though the exported APIs support user space packet access they lack the ability to be invoked in a efficient manner by the application. The lack of support for interacting with these APIs in order to read and write data in parallel from the different process or threads adds to this problem. Moreover since the application is responsible for handling network packets it has to perform the additional operations that would otherwise be carried out by the operating system.

The framework provides support for sending and receiving data in parallel over the network interface from user space and a common parallel execution engine that supports both the application and packet processing tasks.

\subsection{Parallel Packet I/O}
\label{ppio}
The parallel packet I/O interface when combined with the application tasks provides concurrent communication by the application threads or processes over the network. To do this two important requirements are to be met. Firstly, it should provice access to network data from user space which we can write to in parallel and secondly, to integrate it with the application by providing necessary interfaces for performing the packet operations.

\subsubsection{Userspace Packet Access}
Applications executed in the kernel space for accessing hardware devices directly would lead to instability of the entire system. Instead, several previous work \cite{Rizzo:2012:RNI:2090147.2103536}, \cite{1564468}, \cite{Krasnyansky}, etc., propose access of network buffers from user space. We use \texttt{netmap} APIs \cite{Rizzo:2012:RNI:2090147.2103536} to access data from the network buffers. \texttt{Netmap} uses a set of ring buffers in the kernel space and user space to export the network hardware buffer's memory region. This enables the applications to execute in the user space and directly access the network hardware. It prevents the application from accessing memory outside the valid region by checking limits of the ring buffer. Packets can be written to the buffers when the space is available and an \textit{ioctl} call to the driver would enable the network controller to send the data. A \textit{poll} interface is implemented on the memory region mapped in the user space to check when new packets are available. On reading or writing the packets the available space is updated.

We leverage on the availability of multiple hardware queues present in modern NIC to read and write data concurrently to the hardware. \texttt{Netmap} by default supports assigning each hardware queue to an individual application process or thread, but we noticed that in order to be efficient in handling the interface we had to assign more than one queue to a particular thread or process. We modified \texttt{netmap}'s hardware queue assignment API to support this. Later, it also allowed the framework to balance the amount of packets handled by each process by varying the number of queues assigned to each of it in order to balance the workload.

We also found that it was essential to have a balanced control over issuing the system calls with respect to the amount of data being transferred. Frequent system calls seem to hamper the bandwidth and latency and in essence affect the peak performance of the application. To tackle this we propose the following algorithms for reading and writing data from the hardware queues using the \texttt{Netmap} APIs. The algorithms are designed to provide these features,

\begin{itemize}

\item Provide batching of packets based on available space on hardware rings
\item Reduce the usage of system calls such as \textit{ioctl \& poll}
\item Mechanisms for handling multiple queues through the application

\end{itemize}

By minimizing the synchronizations with the hardware rings we essentially limit the system calls issued and improve the overall system performance.

\begin{algorithm}
\caption{To send packets}
\label{s_pack}
\begin{algorithmic}[1]

\State \textbf{struct} $ring[ n ]$\textbf{\{}
\State \Comment{Stores information about hardware rings}
\State \Comment{ $ n \gets \ $tx\ rings\ assigned\ to\ thread}
\State \  \  $avb$, available space
\State \  \  $curr$, current position
\State \  \  $limit$, total space \ \textbf{\}}

\Function{send\_data\_mqueue}{Ring $f(n)$, Data $packet$}
\While{$!success$}
	\State $i \gets f( n ) $ \label{s_pack_param}
	\State $success = $ \texttt{SEND\_DATA}($ i, packet $)
\EndWhile
\EndFunction
\Function{send\_data}{Ring $i$, Data $packet$}
\If{ $ring[ i ].avb > 0$ }
	\State $hw\_ring[ i ].slot( ring[ i ].curr ) \gets packet$
	\State $ring[ n ].avb \gets ring[ i ].avb - 1$
	\State $ring[ i ].curr++$
	\If{ $ring[ i ].avb == 0$ }
		\State $hw\_ring[ i ].avb \ -=\ ring[ i ].limit$
		\State $hw\_ring[ i ].curr \gets ring[ i ].curr$
	\EndIf
	\State \textbf{return} $success$
\EndIf

\State \textbf{\textit{issue ioctl and poll}}
\State $ring[ i ].avb \gets hw\_ring[ i ].avb$
\State $ring[ i ].limit \gets hw\_ring[ i ].avb$
\EndFunction

\end{algorithmic}
\end{algorithm}

Algorithm \ref{s_pack} \& \ref{r_pack} uses a batch update for sending and receiving packets based on the rate at which data flows through the interface. This is done by maintaining a data structure that has the information about the hardware rings and updating it on a periodic basis. Once a fixed amount of space is determined we update this on our local structure, we then continue to read or write data from the hardware ring but update the actual limits and position on our local structure. When we have used up the space that was allocated we then update the hardware ring with the changes done. In cases where the application is generating sparse traffic it would be possible to frequently update the ring in order to maintain a consistent latency in transmission. Multi queue support is provided by managing the process of selecting the ring to send or receive packets based on a application determined function. The $f(n)$ present in Algorithm \ref{s_pack} line \ref{s_pack_param} and Algorithm \ref{r_pack} line \ref{r_pack_param} represents this. Choosing a random queue usually proves to be effective in our experiments but if the application requires a specific flow to be maintained a different function here could provide that support. Cases where this would be helpful is if the data operations are more predictable or if the transactions seem to follow a specific pattern. The $threshold$ specified in Algorithm \ref{r_pack} line \ref{r_pack_thres} is determined for each application based on the rate at which it is able to process data. By issuing the system calls such as \textit{ioctl} and \textit{poll} only when required and effectively handling the multiple queues we are able to achieve improved performance.

\begin{algorithm}[h]
\caption{To receive packets}
\label{r_pack}
\begin{algorithmic}[1]

\State \textbf{struct} $ring[ n ]$\textbf{\{}
\State \Comment{Stores information about hardware rings}
\State \Comment{ $ n \gets \ $rx\ rings\ assigned\ to\ thread}
\State \  \  $avb$, available packets
\State \  \  $curr$, current position
\State \  \  $limit$, total packets \ \textbf{\}}

\Function{receive\_data\_mqueue}{Ring $f(n)$} 
\While{$!packet$}
	\State $i \gets f( n ) $ \label{r_pack_param}
	\State $packet =$ \texttt{RECEIVE\_DATA}(Ring $i$)
\EndWhile
\EndFunction
